Automatic extraction of symbolic network signatures from malware code

### 2016-10-31 paper.tex
add related work

\section{Project Proposal}
\textbf{Tasks}:
\begin{itemize}
	\item Identify and characterize features of common malware code - 1-2 weeks
	\added{Refer to ---}
	\item \deleted{Develop the abstract semantics and operators for extracting  
	symbolic network signatures from malware code - 2-3 weeks} \\
	\added{Automatically generate a symbolic automaton of the given program
		based on packet and variable state - 2-3 weeks}
	\item Modifying/Extending existing tools like Codesurfer for symbolic 
	network signature extraction - 2-3 weeks
\end{itemize}

@@

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] = "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] = "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] = "c") {
		attack()  [the attack point is provided as input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
The symbolic automaton for this program is 
shown in \Cref{fig:progautomaton}. Network signatures
for the malware are all paths from start state ($a=0$)
to state of attack. In this case, the sequence of packets
in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
\begin{figure}[h]
	\centering
	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
	\node[state, initial] (v0)   {$a = 0$}; 
	\node[state] (v1) [right=of v0] {$a = 1$}; 
	\node[state] (v2) [right=of v1] {$a = 2$};
	\node[state] (v3) [right=of v2] {$attack$}; 	
	\path[->] 
	(v0) edge node {$a*$} (v1)
	(v1) edge node {$b*$} (v2)
	(v2) edge node {$c*$} (v3);
	\end{tikzpicture}
	\caption{Symbolic automaton for malware program.}
	\label{fig:progautomaton}
\end{figure}
\bibliographystyle{abbrv}
\bibliography{references}

>>>>

\section{Project Proposal}
\textbf{Tasks}:
\begin{itemize}
	\item Identify and characterize features of common malware code - 1-2 weeks
	\added{Refer to sections A and B}
	\item \deleted{Develop the abstract semantics and operators for extracting  
	symbolic network signatures from malware code - 2-3 weeks} \\
	\added{Automatically generate a symbolic automaton of the given program
		based on packet and variable state - 2-3 weeks}
	\item Modifying/Extending existing tools like Codesurfer for symbolic 
	network signature extraction - 2-3 weeks
\end{itemize}

@@

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] = "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] = "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] = "c") {
		attack()  [the attack point is provided as 
				input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
The symbolic automaton for this program is 
shown in \Cref{fig:progautomaton}. Network signatures
for the malware are all paths from start state ($a=0$)
to state of attack. In this case, the sequence of packets
in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
\begin{figure}[h]
	\centering
	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
	\node[state, initial] (v0)   {$a = 0$}; 
	\node[state] (v1) [right=of v0] {$a = 1$}; 
	\node[state] (v2) [right=of v1] {$a = 2$};
	\node[state] (v3) [right=of v2] {$attack$}; 	
	\path[->] 
	(v0) edge node {$a*$} (v1)
	(v1) edge node {$b*$} (v2)
	(v2) edge node {$c*$} (v3);
	\end{tikzpicture}
	\caption{Symbolic automaton for malware program.}
	\label{fig:progautomaton}
\end{figure}

\section{Related Work}
An important assumption that we make is that of the simplification of the botnet client code to a set of \textbf {if (state, packet)} statements. Looking at existing research, significant work has gone into automatic protocol reverse-engineering of botnet C\&C protocols using dynamic binary analysis based on network traces. This generally follows a two step process: Extract the protocol grammar, that captures the structure of messages comprising the protocol. Knowing the protocol message format would enable the construction of the protocol state machine, which represents the sequence of messages as a specification of protocol states and transition between states (based on messages received). 
For example, Polyglot~\cite{polyglot} is a system that uses dynamic binary analysis to automatically extract the protocol message format, and Dispatcher~\cite{dispatcher} is a tool that makes use of techniques specified in Polyglot to extract the message format and field semantics for messages sent and received by the application.

Taint propagation is used by the tool to map how the packets received manipulate the state. When the program calls the \textit{read()} function, Dispatcher uses the function's prototype to determine if any argument is tainted based on the memory locations it accesses. Via this approach, a spam botnet called MegaD is analyzed by Dispatcher.

Thus, having knowledge of the network packets received would enable us to make inferences on how they could transition to an 'attack point'. Our work focuses on static analysis of the malware code, without having to execute it and/or capture network traces.

In addition, work has been done on inferring state machines automatically for stateful network protocols - Prospex~\cite{prospex}. Prospex essentially performs clustering of similar network messages and tracks protocol state changes. By merging similar states, a minimal state machine is produced, and the paper produces a labeled state tree for the Agobot malware. This could be used as our initial malware program model for which we determine the sequence of packets to initiate an attack.

\bibliographystyle{abbrv}
\bibliography{references}
